---
title: Financial data aggregator
author: Brent Morrison
date: '2018-10-21'
slug: financial-data-aggregator
categories:
  - Data
tags: []
description: ''
topics: []
---

This post is going to document an R script that will download and aggregate various financial and economic time series.  This data will be used in future posts to perform modelling and analyse.  

### Packages required and download parameters


Load the required packages.
```{r code_chunk_1, error=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
library("DescTools")
library("tidyquant")
library("timetk")
library("broom")
library("tibbletime")
```


We then set parameters specifying the data to be downloaded.
```{r code_chunk_2, error=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
# Parameters for binary market indicator
lb = 6                  #  Lookback period
pc = 0.2                #  Percent drawdown for binary market in/out indicator cutoff
fr = -0.025             #  Forward return for binary market in/out indicator cutoff

# Start and end dates
s.date = as.Date("1945-01-01")
e.date = as.Date("2018-11-01")

# Time series to download, Quandl and FRED
qndlm = c("NAHB/NWFHMI.1",
          "ISM/MAN_NEWORDERS.5")
fredw = c("IC4WSA")
fredm = c("AAA",       #	Moody's Seasoned Aaa Corporate Bond Yield
          "ACDGNO",    #	Value of Manufacturers' New Orders for Consumer Goods: Consumer
          "AHETPI",    #	Average Hourly Earnings of Production and Nonsupervisory Employees:
          "AWHMAN",    #	Average Weekly Hours of Production and Nonsupervisory Employees:
          "BAA",       #	Moody's Seasoned Baa Corporate Bond Yield
          "BOGMBASE",  #	Monetary Base; Total
          "CFNAIDIFF", #	Chicago Fed National Activity Index: Diffusion Index
          "CPIAUCSL",  #	Consumer Price Index for All Urban Consumers: All Items
          "CPILFESL",  #	Consumer Price Index for All Urban Consumers: All Items Less Food and
          "FEDFUNDS",  #	Effective Federal Funds Rate
          "GS10",      #	10-Year Treasury Constant Maturity Rate
          "GS2",       #	2-Year Treasury Constant Maturity Rate
          "INDPRO",    #	Industrial Production Index
          "ISRATIO",   #	Total Business: Inventories to Sales Ratio
          "KCFSI",     #	Kansas City Financial Stress Index
          "M2SL",      #	M2 Money Stock
          "NEWORDER",  #	Manufacturers' New Orders: Nondefense Capital Goods Excluding Aircraft
          "PERMIT",    #	New Private Housing Units Authorized by Building Permits
          "TB3MS",     #	3-Month Treasury Bill: Secondary Market Rate
          "TWEXMMTH",  #  Trade Weighted U.S. Dollar Index: Major Currencies
          "UNRATE",    #	Civilian Unemployment Rate
          "LOANS")     #	Loans and Leases in Bank Credit, All Commercial Banks 
```

```{r code_chunk_3, error=FALSE, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
quandl_api_key("hpbPcsfGudN3viBgh8th")
```

###Downloading external data


This section of code documents the actual capture of data from various sources.  We are using the "tq_get" function from the tidyquant package.

```{r code_chunk_4, error=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
# Get S&P500 data
sp_5   <-  tq_get("^GSPC",get = "stock.prices",from = s.date)

# Get fred monthly economic data
econ.m1 <- tq_get(fredm, get = "economic.data",from = s.date)

# Spread fred monthly data to column, fill missing values
econ.m2 <- spread(econ.m1, symbol, price) %>% 
  fill(ACDGNO, CFNAIDIFF, ISRATIO)

# Get fred weekly economic data
econ.w1 <- tq_get(fredw, get = "economic.data", from = s.date)

# Convert weekly data to monthly frequency
econ.m3 <- econ.w1 %>% 
  rename("IC4WSA" = "price") %>%
  group_by(month=floor_date(date, "month")) %>% 
  summarize(IC4WSA = last(IC4WSA)) %>%
  rename("date" = "month") 

# Get quandl monthy data
econ.m4 <- tq_get(qndlm,get="quandl",from="1985-03-01") %>%
  mutate(price = if_else(is.na(value), index, value), 
         date = floor_date(if_else(is.na(date), month, date),"month")) %>%
  select(symbol, date, price) %>% spread(symbol, price) %>%
  rename(HMI = "NAHB/NWFHMI.1", NEWORD = "ISM/MAN_NEWORDERS.5") 

# Get Shiller download: http://www.econ.yale.edu/~shiller/data.htm
econ.m5 <- read.zoo(file = "C:/Users/brent/Documents/R/R_import/Shiller.csv",
                    FUN = as.Date, header = T, sep = ",", 
                    format = "%d/%m/%Y", index.column = 1)
econ.m5 <- tk_tbl(econ.m5, rename_index = "date") %>% 
  mutate(date = floor_date(date, "month"))

# Join all data (except stock data)
econ.m  <- full_join(econ.m2, econ.m3, by = "date")
econ.m  <- full_join(econ.m, econ.m4, by = "date")
econ.m  <- full_join(econ.m, econ.m5, by = "date") %>% 
  filter(date >= s.date & date <= e.date)
```
Note that the Shiller cyclically adjusted price earnings data is manually saved down and loaded via a read csv function.

Let's see what our data looks like at this stage.
```{r code_chunk_5, error=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
str(econ.m)
```
Obviously not all times series start at the specified start date.  Everything requested is there, however we have not joined the S&P 500 index data as yet.

### Stock market data

We now want to take the stock data (the S&P 500), roll up from a daily to monthly periodicity, and create a number of rolling attributes
```{r code_chunk_6, error=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
# Create new attributes
sp_5 <- sp_5 %>% 
  
  # Group to monthly periodicity and create monthly returns
  group_by(month = floor_date(date, "month")) %>%
  summarize(low = min(low), close = last(close), volume = sum(volume)) %>%
  rename("date" = "month") %>% 
  tq_mutate(select = close, mutate_fun = periodReturn, period = "monthly", 
            type = "log", col_rename  = "rtn_m") %>%
  
  # Forward monthly return
  mutate(fwd_rtn_m = lead(rtn_m, 1)) %>% 
  
  # 6 monthly returns
  tq_mutate(select = rtn_m, mutate_fun = rollapply, width = lb, FUN = sum, 
            col_rename  = "rtn_6m") %>%
  
  # 6 monthly low watermark
  tq_mutate(select = low, mutate_fun = rollapply, width = lb, FUN = min, 
            col_rename  = "min_6m") %>%
  
  # 6 monthly max drawdown
  mutate(dd_6m = -lag(log(close), n = lb) + log(min_6m)) %>%
  
  # Binary flag, 1 = 6 monthly return less than specified and drawdown in
  # in excess ofthat specified 
  mutate(flag = ifelse(rtn_6m < fr | dd_6m < -pc , 1, 0)) %>%
  
  # Lead binary flag by 6 months
  mutate(y1 = lead(flag, lb)) %>%
  
  # Indicator for when binary flag changes
  mutate(diff_flag = c(NA, diff(y1)))
```

The attribute y1 is a binary indicator that looks forward over the next 6 months and return 1 if the maximum drawdown is greater than 20% or the return is less than 2.5% (these parameters can be changed as detailed above).  This will be the independent or forecast variable for future analysis.

If we can forecast this indicator we can stay out of the market during downturns or when returns are deemed insufficient.

In order to visualise this indicator it is ideal to do so as a shading over a time series chart. In order to do that we need to convert the indicator into a series of start and end dates, the following code does just that.
```{r code_chunk_7, error=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
sp_5s    <-  sp_5 %>% filter(diff_flag == 1) %>% select(date) %>% rename(start = date)
sp_5e    <-  sp_5 %>% filter(diff_flag == -1) %>% select(date) %>% rename(end = date)
short    <-  min(count(sp_5s), count(sp_5e))
sp_shade <-  data.frame(head(sp_5s, short), head(sp_5e, short))
```

All data sets have now been created.  Lastly these data sets are joined and saved down for future reference.
```{r code_chunk_8, error=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
# Join economic and stock return time series
econ_fin_data <- inner_join(econ.m, sp_5, by = "date")
# Save time series data and 
saveRDS(econ_fin_data, file="econ_fin_data.Rda")
saveRDS(sp_shade, file="sp_shade.Rda")
```
That's it for now, next we will look at visualising these time series.