---
title: Kable correction
author: Brent Morrison
date: '2018-12-16'
slug: kable-tables
categories:
  - post
tags:
  - test
description: ''
topics: []
---

Lets try and resolve the issue I reported [here](https://brentmorrison.netlify.com/post/test-post/) . This code attempts to rectify the formatting issue by closing off the code chunk appropriately.   I've also loaded the tidyverse packages as was done in this [stackoverflow](https://stackoverflow.com/questions/48409288/how-to-show-kable-kableextra-tables-in-r-markdown-notebook-not-just-in-the-knit?rq=1) post.

```{r code_chunk_1, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
  library(kableExtra)
  library(tidyverse)
  library(knitr)
  dt <- mtcars[1:15, 1:6]
  kable(dt, format = "html", align = "c") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), 
                  full_width = T,
                  position = "left") %>%
    column_spec(1, width = "10em") %>% 
    column_spec(2, width = "7em") %>% 
    column_spec(3, width = "7em") %>% 
    column_spec(4, width = "7em") %>% 
    column_spec(5, width = "7em") %>% 
    column_spec(6, width = "7em") %>% 
    footnote(general = "Here is a general comments of the table. ",
             number = c("Footnote 1; ", "Footnote 2; "),
             alphabet = c("Footnote A; ", "Footnote B; "),
             symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"))
```

That does not have any impact.  

Let's try something different.
```{r code_chunk_2, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
kable(dt) %>%
  kable_styling("striped") %>%
  add_header_above(c(" " = 1, "Group 1" = 2, "Group 2" = 2, "Group 3" = 2))
```

Again, the format differs to that detailed in the vignette, no horizontal lines are produced.   

Lastly, let's try xtable using the image linked in this [stackoverflow](https://stackoverflow.com/questions/37449067/use-xtable-to-print-html-table-in-r-markdown-file) thread.

```{r code_chunk_3, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(xtable)
print(xtable(dt), type = "html")
```
No go.  It looks like I'll have to come back to this one....

