---
title: IFRS9 disclosures (part 3)
author: Brent Morrison
date: '2020-01-18'
slug: ifrs9-disclosures-part-3
categories:
  - Accounting
tags:
  - R
  - Accounting
description: ''
topics: []
---

So we continue on our IFRS9 disclosures quest!  

[Part 2](https://brentmorrison.netlify.app/post/ifrs9-disclosures-part-2/) had us doing some heavy data munging, followed by modeling to get an ECL balance.  

In this post we will massage this dataset into the report we set out to build in the first post.

```{r opts, include=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE, echo=TRUE, fig.width=8.25, fig.height=8.25)
```

```{r packages}
library("tidyverse")
library("DescTools")
library("lubridate")
library("tibbletime")
library("scales")
library("kableExtra")
library("DT")
library("widgetframe")
library("htmltools")
```

The obvious starting point is to grab the data from part 2.  This data was saved as both a `feather` and `rds` file.  These are online in the blogs repo [here](https://github.com/Brent-Morrison/hugo_website/tree/master/content/post).  I'm going to load the local copy for now. 

```{r data}
ifrs9_part2 <- readRDS(file = "ifrs9_part2.Rds")
```

Below I transform attribute names to a format based on a separate implementation of this dataflow.  Additional attributes have been generated here:  
  * Currency `ccy`, the (imaginary) currency of the loan  
  * Type `type`, this assigns a stock (a customer of our bank) as either revolving or term.  A revolving exposure can be repaid and subsequently drawn upon again in contrast to term loans which are drawn down once  
  * "Purchased or originated financial asset(s) that are credit-impaired on initial recognition", I won't go into that mouthful here. Just know that it is a required classification  
  * The ledger balance of the loan `bal`. This will differ to the gross carrying amount if fees paid or received on orgination have been deferred  
  * `wof`, write-offs  
  * `pryr`, prior year balance for stage 1 loans assuming no model input changes   
  * `prlt`, prior year balance for stage 2 and 3 loans assuming no model input changes  

This task has been worked on separately [here](https://github.com/Brent-Morrison/Financial_reporting/blob/master/Fin_reporting.R). To take advantage of this work I need to rename variables.

```{r transform}
ifrs9_part3 <- ifrs9_part2 %>% 
  rename(
    cust = Ticker,
    unit = clust.name,
    stage = RiskStage,
    gca = TotalDebt,
    ecl = ECL
    ) %>% 
  mutate(
    date = me.date,
    # Assign missing attributes
    ccy = if_else(str_detect(cust, "^G"), "GBP", "USD"),
    type = if_else(str_detect(cust, "^R"), "rvlv", "term"),
    poci = if_else(str_detect(cust, "^P") & stage == 3, "Y", "N"),
    bal = gca,
    ecl = -ecl,
    wof = 0,
    pryr = 0,
    prlt = 0
  ) %>% 
  select(-me.date)
```

Here we go with the mother of all transforms....

```{r create_tables_1}
#=========================================================================================
#==    Create stage attributes & movement balances                                      ==
#==    1. assign disclosure stage (1 / 2 / 3 / 4=POCI_NCI / 5=POCICI)                   ==
#==    2. fill stage with preceding value (ignores initial na's)                        ==
#==       & default remaining na's to stage 1                                           ==
#==    3. create lagged values of balances and prior period balance                     ==
#==    4. rename closing balance attributes                                             ==
#==    5. add cumulative movement                                                       ==
#==    TO DO - add FINREP change of stage indicator for table 12.1                      ==
#=========================================================================================

#min.date <- ifrs9_part3 %>% slice(which.min(date)) %>% select(date)

bln_ctgy <- ifrs9_part3 %>%  
  mutate(ctgy = case_when(
    wof != 0     & poci == "N" ~ 3,                                                     #1
    wof != 0     & poci == "Y" ~ 5,
    wof != 0     & is.na(poci) ~ 3,
    stage == 1   & poci == "N" ~ 1,
    stage == 2   & poci == "N" ~ 2,
    stage == 3   & poci == "N" ~ 3,
    is.na(stage) & poci == "N" ~ 1,
    stage == 1   & poci == "Y" ~ 4,
    stage == 2   & poci == "Y" ~ 4,
    stage == 3   & poci == "Y" ~ 5,
    is.na(stage) & poci == "Y" ~ 4,
    stage == 1   & is.na(poci) ~ 1,
    stage == 2   & is.na(poci) ~ 2,
    stage == 3   & is.na(poci) ~ 3)
  ) %>% 
  group_by(cust) %>% fill(ctgy) %>% replace_na(list(ctgy = 1)) %>%                      #2
  mutate(gca.op = lag(gca),                                                             #3        
         bal.op = lag(bal),
         bal.pr = if_else(date == which.min(date), bal, 0),
         ecl.op = lag(ecl),
         ctgy.op = lag(ctgy)) %>% 
  fill(bal.pr) %>% 
  rename(gca.cl = gca,                                                                  #4
         bal.cl = bal, 
         ecl.cl = ecl,
         ctgy.cl = ctgy, 
         wof.cl = wof
         ) %>%                                            
  mutate(bal.y     = bal.cl - bal.pr + cumsum(wof.cl),                                  #5
         bal.y.dd  = if_else(bal.y > 0, bal.y, 0),
         bal.y.rd  = if_else(bal.y < 0, bal.y, 0)) %>% 
  ungroup()


#=========================================================================================
#==    Create movement attributes                                                       ==
#=========================================================================================

bln_mvnt <- bln_ctgy %>% group_by(cust) %>% 
  mutate(cover.cl     = -ecl.cl / bal.cl,
         cover.op     = -ecl.op / bal.op,
         cover        = Winsorize(round(if_else(is.nan(cover.op),cover.cl, cover.op), 5), 
                                  minval = 0, maxval = 1),
         incr.decr    = case_when(bal.cl > bal.op ~ 'incr',
                              bal.cl < bal.op ~ 'decr',
                              TRUE ~ 'unch'),
         ctgy.dir     = case_when(ctgy.cl > ctgy.op ~ 'D',
                              ctgy.cl < ctgy.op ~ 'I',
                              TRUE ~ 'U'),
         pre.post     = case_when(ctgy.dir == 'I' & incr.decr == 'decr' ~ 'pre',
                              ctgy.dir == 'D' & incr.decr == 'incr' ~ 'pre',
                              TRUE ~ 'post'),
         pre.stage    = if_else(pre.post == 'pre', ctgy.op, ctgy.cl),
         gca.m.dd.r   = if_else(type == 'rvlv', bal.y.dd - lag(bal.y.dd), 0),
         gca.m.dd.t   = if_else(type == 'term' & incr.decr == 'incr', 
                                bal.cl - bal.op + wof.cl, 0),
         gca.m.rd.t.f = if_else(type == 'term' & incr.decr == 'decr' & bal.cl == 0, 
                                bal.cl - bal.op + wof.cl, 0),
         gca.m.rd.t   = if_else(type == 'term' & incr.decr == 'decr' & bal.cl != 0, 
                                bal.cl - bal.op + wof.cl, 0),
         gca.m.rd.r   = if_else(type == 'rvlv', bal.y.rd - lag(bal.y.rd), 0),
         gca.m.oth    = (gca.cl - bal.cl) - (gca.op - bal.op),
         g.tfr.pre    = gca.op + gca.m.dd.r + gca.m.dd.t + gca.m.rd.t.f + 
                        gca.m.rd.t + gca.m.rd.r,
         gca.m.wof    = -wof.cl,
         gca.m.tfr.o  = -case_when(ctgy.dir != 'U' & pre.post == 'pre' ~ g.tfr.pre,
                                   ctgy.dir != 'U' & pre.post == 'post' ~ gca.op,
                                   TRUE ~ 0),
         gca.m.tfr.i  = -gca.m.tfr.o,
         ecl.m.dd.r   = -cover * gca.m.dd.r,
         ecl.m.dd.t   = -cover * gca.m.dd.t,
         ecl.m.rd.t.f = -cover * gca.m.rd.t.f,
         ecl.m.rd.t   = -cover * gca.m.rd.t,
         ecl.m.rd.r   = -cover * gca.m.rd.r,
         ecl.m.wof    = wof.cl,
         ecl.m.prm    = case_when(ctgy.cl == 1 & pryr != 0 ~ ecl.cl + pryr,
                                  ctgy.cl != 1 & prlt != 0 ~ ecl.cl + prlt,
                                  TRUE ~ 0),
         ecl.m.rem.mig= if_else(ctgy.dir != 'U', ecl.cl - ecl.op - ecl.m.dd.r - 
                                                 ecl.m.dd.t - ecl.m.rd.t.f - 
                                                 ecl.m.rd.t - ecl.m.rd.r - 
                                                 ecl.m.wof - ecl.m.prm, 0),
         ecl.m.rem    = if_else(ctgy.dir == 'U', ecl.cl - ecl.op - ecl.m.dd.r - 
                                                 ecl.m.dd.t - ecl.m.rd.t.f - 
                                                 ecl.m.rd.t - ecl.m.rd.r - 
                                                 ecl.m.wof - ecl.m.prm, 0),
         ecl.tfr.pre  = ecl.op + ecl.m.dd.r + ecl.m.dd.t + 
                        ecl.m.rd.t.f + ecl.m.rd.t + ecl.m.rd.r,
         ecl.m.tfr.o  = -case_when(ctgy.dir != 'U' & pre.post == 'pre' ~ ecl.tfr.pre,
                                   ctgy.dir != 'U' & pre.post == 'post' ~ ecl.op,
                                   TRUE ~ 0),
         ecl.m.tfr.i = -ecl.m.tfr.o) %>% ungroup() %>% 
  select(-g.tfr.pre, -ecl.tfr.pre) %>% 
  mutate_at(vars(starts_with("g.")), list(~replace_na(., 0))) %>%  
  mutate_at(vars(starts_with("i.")), list(~replace_na(., 0))) %>%
  mutate(gca.ch  = gca.op + rowSums(select(., contains("gca."))) - gca.cl,
         ecl.ch  = ecl.op + rowSums(select(., contains("ecl."))) - ecl.cl)


#=========================================================================================
#==     Gather to long table  & apply pre / post rules                                  ==
#=========================================================================================

bln_mvnt_long <- bln_mvnt %>% 
  select(date, cust, ctgy.cl, ctgy.op, pre.post, gca.cl, 
         ecl.cl, gca.op, ecl.op, gca.m.dd.r:ecl.m.tfr.i) %>% 
  gather(key = "m.ment", value = "tran_ccy", gca.cl:ecl.m.tfr.i) %>% 
  arrange(cust, m.ment, date) %>% filter(tran_ccy != 0) %>% 
  mutate(bal.type = str_sub(m.ment, 0, 3),
         m.type   = str_sub(m.ment, 5)) %>% 
  mutate(ctgy = case_when(m.type == "op" |
                          m.type == "m.tfr.o" |
                          m.type == "m.dd.r"   & pre.post == "pre" |
                          m.type == "m.dd.t"   & pre.post == "pre" |
                          m.type == "m.rd.t.f" & pre.post == "pre" |
                          m.type == "m.rd.t"   & pre.post == "pre" |
                          m.type == "m.rd.r"   & pre.post == "pre" ~
                          ctgy.op,
                          TRUE ~ ctgy.cl), 
         m.type = if_else(str_detect(m.type, "tfr"), 
                          paste("tfr", ctgy.op, ctgy.cl, sep = "."), 
                          m.type),
         year = year(date),
         month = month(date)) %>% 
  select(date, year, month, cust, ctgy.op, ctgy.cl, ctgy, bal.type, m.type, tran_ccy) %>% 
  arrange(cust, date, bal.type)

# TO DO: unsure no duplicates over account/company/month

# Save for Tableau
#saveRDS(as.data.frame(bln_mvnt_long), file = "bln_mvnt_long.rda")
#write.csv(bln_mvnt_long, file = "bln_mvnt_long.csv")
```


```{r create_tables_2}
report <- bln_mvnt_long %>% 
  filter(
    cust == "ORCL",
    date == "2017-06-30"
    ) %>% 
  select(
    -date:-ctgy.cl
    ) %>% 
  pivot_wider(
    names_from = c(bal.type, ctgy),
    values_from = tran_ccy
    ) 
```